Supplementary Information: Tropical tree drought sensitivity is jointly shaped by drought characteristics, species adaptations, and individual microenvironments

### 2025-06-19 doc/manuscript_hkk_drought_sensitivity.Rmd
some more comments from HML

<!--PARAGRAPH ON DROUGHT CHARACTERISTICS-->
<!-- Sean::: This paragraph is a bit redundant and all over the place. 1) in the tropics, seasonal dry periods (often defined as months with <= 100 mm rainfall) can have complex influences on tree phenology, from leaf turnover to growth allocation. Droughts, which can exacerbate or extend a dry season, or reduce cloud cover ina rainy season, would be expected interact with seasonal phenology also in complex ways. [Then go into some details with citations on ?] -->
Drought is notoriously hard to define and encompasses a variety of types based on the meteorological variables affected and the timing, severity and duration of climatic anomalies [@slette_how_2019] with potentially distinct ecological consequences.
<!-- ; we define it here as hot/dry conditions anomalously departing from climatic means during any time of year  -->
Although forest drought responses are often considered to follow consistent patterns across droughts [e.g.,@bennett_larger_2015; @brando_droughts_2019], droughts with different timing, duration, or intensity may have distinct effects on tree growth, especially in seasonal forests with adaptive phenological strategies to handle regular periods of climatic stress [@albert_cryptic_2019] [@gao_dynamic_2018].
<!-- Tropical tree growth shows predictable seasonal patterns [@garcía-cervigón_climate_2020;@giraldo_tree_2023]; drought timing can interact with this growth phenology especially.  -->
<!-- Sean::: don't think the above is true universally. Laura Boeschoten and Erick Esteban have separate papers well in prep showing some species demonstarating seasonal growth and others not -->
In tropical forests with dry seasons unfavorable to growth, tree drought responses may vary dramatically depending on the timing of drought relative to seasonal cycles of climate, photosynthesis, and woody growth [@garcía-cervigón_climate_2020]. <!---- agreed but see above ref I put in to [@albert_ecology_letters_cryptic_phenology]. It seems like these arguments are a bit redundant. Or meandering. ---->
<!--(Here we distinguish dry season from drought; dry season represents hot/dry conditions that are periodic, annual and predictable, droughts are unpredictable eith anomalous hot, dry conditions relative to seasonal expectations.) -->
Analyses to date have found that tropical tree growth <!--decreases with temperature[@zuidema_tropical_2022;@clark_annual_2010;@clark_annual_2021; @vlam_temperature_2014] and -->is particularly sensitive to dry season precipitation [@zuidema_tropical_2022;@clark_annual_2010;@clark_annual_2021]; however, these analyses represent a very limited set of tropical tree species and climates. 
It remains unknown how diverse assemblages of tropical tree species <!--in tropical dry forests -->respond to different types of droughts, including those that intensify or conflict with regular seasonal cycles<!-- of wet and dry seasons-->. 
<!--
Seasonally dry tropical forest climates are often close to thermal and water stress tolerance thresholds for many species therein, making these forests vulnerable to drought [@allen_will_2017] and leading to stand-level patterns of drought-associated water stress [@bauman_tropical_2022; @bennett_sensitivity_2023; @aguirre-gutiérrez_drier_2019]. <!--the last phrase of this sentence and how it relates to Bennett et al. is unclear to me.
Trees in seasonally dry tropical forests exhibit a mix of strategies - along a spectrum of conservative to acquisitive water-use - that support survival and growth despite strong seasonal hot/dry stress.
These strategies might confer differential advantages under drought conditions. -->

@@

<!-- Sean::: mention rooting depth? [@chitra-tarak: root of the drought and chitra-tarak: new_phyt 2022] -->
Tree species adaptations to drought vary widely in type and strength, often leading to classification of species as "drought tolerant" or "drought sensitive" [@guillemot_small_2022; @vico_how_2017a; @oliveira_linking_2021].
In tropical seasonal forests in particular, trees have evolved to maximise growth, reproduction, and survival under dry seasons, with strategies ranging along a spectrum from hydraulic safety to efficiency, leading to tolerance or avoidance of dry conditions [@gonzález-m._diverging_2021].
Dry-season deciduousness represents an extreme "avoidance" strategy along this continuum, where species lose leaves during the dry season to avoid foliar water loss and have acquisitive strategies during leaf-on periods [@desouza_drought_2020]. 
Other strategies include deep roots (hydraulic safety, to access deeper water in the dry season), more negative leaf water potential at turgor loss point ($\pi_{tlp}$, hydraulic safety, to maintain gas exchange and photosynthesis under hot and dry conditions), and large vessels (hydraulic efficiency)<!--but this may put them at risk of embolism--> [@gonzález-m._diverging_2021;@chitra-tarak_hydraulicallyvulnerable_2021]. 
Under drought conditions, some of these strategies can lend drought resistance while others can amplify drought sensitivity, especially under severe droughts [@kunert_leaf_2021; @chitra-tarak_hydraulicallyvulnerable_2021; @gonzález-m._diverging_2021]. 
<!-- This spectrum of species strategies in seasonally dry forests for demographic stability under hot dry seasons might have distinct consequences for sensitivity, based on drought characteristics.  -->
<!--Some theoretical evidence that the relative advantage of evergreen and deciduous strategies might differ with changing precipitation regimes [@vico_how_2017a]--->
Moreover, besides being adaptively conserved, these strategies can also have developmental or ecophysiological controls.
For example, some species regulate negative $\pi_{tlp}$'s, leaf deciduousness, and root depth based on environmental cues, leading to intraspecific or interannual variation and potentially resulting in spatiotemporal variation in drought responses [@williams_deciduousness_2008; @hulshof_variation_2010].

@@

Our study site was the Huai Kha Khaeng ForestGEO plot (15.6324 N, 99.217 E), hereafter "HKK", in the Huai Kha Khaeng Wildlife Sanctuary, Uthai Thani, Thailand.
This is a 50-hectare plot in dry seasonal evergreen forest, with mean annual precipitation of 1400 mm and a strong dry season from November to April [@anderson-teixeira_ctfsforestgeo_2015].
The region is periodically affected by droughts associated with the *El Niño* Southern Oscillation [ENSO, @räsänen_spatial_2016].
The dominant canopy species are _Hopea odorata_, _Vatica harmandiana_, _Dipterocarpus alatus_ and _Tetrameles nudiflora_ and lower layers of the canopy are dominated by sub-canopy evergreen and deciduous species like _Miliusa horsfieldii_ and _Polyalthia viridis_. 
<!-- For analysis, we focused on 30 of the most abundant species in the plot.    -->
<!-- Analysis of long-term tree-ring data shows that interannual variation tree growth at HKK is influenced by temperature and rainfall [@vlam_temperature_2014].   -->

@@

Using the late wet/early dry dendrometer band censuses, we calculated annualised growth for each year and tree. 
We first conducted quality control on the dataset, removing measurements that appeared likely to be data entry errors (misplaced decimals), potential misidentified bands (old band numbers that restarted after stopping) or misidentified individuals (individuals with conflicting metadata across censuses).
We used window size measurements combined with the diameter measurement at installation $(d_{1})$ to calculate diameter at each dendroband census using standard equations that uses the chord across the measurement window to estimate the arc, assuming a circular cross section[@condit_dendrometer_].
We used a direct measurement of $d_{1}$ at the time of installation if it was available; if not, we excluded window size measurements until the first available diameter measurement. 
We calculated annualised diameter increments for each individual $i$ for each year $t$ by finding the difference in diameter from the previous year ($inc_{i,t} = d_{i,t} - d_{i,t-1}$) and adjusting for the number of days ($tdif$) between measurements ($ainc_{i, t} = inc_{i,t} * \frac{tdif}{365}$).

<!-- To ensure that the data did not smooth over multiple years, we excluded increment measures spanning missing values for $d_{t-1}$. -->
We excluded large measurement outliers (> 3 standard deviations from the mean increment across all observations). 
Diameter tape measurements themselves are not precise enough for calculating drought sensitivity; error for each measurement is 1 mm, error for growth is 2 mm. 
In comparison, median annual growth across trees in HKK is ~1.4 mm and calculating deviation from annual growth needs sub-mm precision. 
However, we used tape measurements to flag likely errors in the dendrometer dataset. 
We calculated increments from annual tape measurements made on each dendrobanded tree at each census and removed dendroband measurements that had low agreement. 
Finally, we excluded trees with negative increments over the whole timeseries, and included only species that had at least 10 individuals, resulting in a final dataset of 1820 individuals of 30 species.

@@

Across all individuals, we calculated Topographic Wetness Index (TWI) for an expectation of the water availability across space based on the total upslope area and the slope at the location. 
We used a Digital Elevation Model from a 5 m x 5 m elevation layer collected at plot establishment and the *whitebox* package in R [@wu_whitebox_2022; @lindsay_whitebox_2016a] to calculate flow accumulation at each cell and then calculate TWI as $log_{e}\frac{As}{tan(slope)}$.
<!-- As an alternate measure of water availability, we also calculated Topographic Position Index (TPI) based on the concavity/convexity of a location.  -->
For an alternate estimate of wetness (Topographic Position Index) and associated results, see supplementary material. 
We defined tree size as the diameter at breast height (calculated from dendrometer band window measurement) at the previous census. 
We centred and scaled TWI and DBH across the range of observed values. 
We estimated crown exposure for each tree and year using ground measurements of crown illumination index (CII), an ordinal variable from 1 to 5 (least to most exposed).

We assigned deciduousness as a species-level trait using data from @williams_deciduousness_2008. 
We defined deciduousness as the mean over trees of the proportion of crown leaf loss at maximum, with a score ranging from 0 to 4 where 0 is most evergreen and 4 is most deciduous. 
Data were unavailable for _Alphonsea ventricosa_, an evergreen species, and so we manually assigned it a deciduousness score of 1. 
We also considered an alternative measure of deciduousness - the duration of time spent at maximum crown loss.
This was strongly correlated with the proportion of crown loss; we chose to use the proportion of crown loss because it was available for more species.  
<!-- At the species level, we also calculated the maximum size across all individuals in the plot as measured in the 2008 whole-plot census.
We calculated niche breadth as the standard deviation of TWI across all observed individuals of each species in the whole plot census in 2008.  -->

@@

We then used the predicted sensitivities from this model and tested their correlations with deciduousness using simple linear models.

@@

We then calculated predictions across all combinations of values from observed maximum to minimum of TWI and deciduousness within this dataset.

To further disentange the microenvironmental drivers of drought sensitivity, we simultaneously modelled:

$$
sensitivity \sim DBH + CII + TWI + (DBH+CII+TWI | Species)
$$ 
and 
$$
CII \sim DBH
$$

>>>>

<!--PARAGRAPH ON DROUGHT CHARACTERISTICS-->
<!-- Sean::: This paragraph is a bit redundant and all over the place. 1) in the tropics, seasonal dry periods (often defined as months with <= 100 mm rainfall) can have complex influences on tree phenology, from leaf turnover to growth allocation. Droughts, which can exacerbate or extend a dry season, or reduce cloud cover ina rainy season, would be expected interact with seasonal phenology also in complex ways. [Then go into some details with citations on ?] -->
Drought is notoriously hard to define and encompasses a variety of types based on the meteorological variables affected and the timing, severity and duration of climatic anomalies [@slette_how_2019] with potentially distinct ecological consequences.
<!-- ; we define it here as hot/dry conditions anomalously departing from climatic means during any time of year  -->
Although forest drought responses are often considered to follow consistent patterns across droughts [e.g.,@bennett_larger_2015; @brando_droughts_2019], droughts with different seasonal timing, duration, or intensity may have distinct effects on tree growth, especially in seasonal forests with diverse adaptive strategies to handle regular dry periods [@albert_cryptic_2019;@gao_dynamic_2018].
<!-- Tropical tree growth shows predictable seasonal patterns [@garcía-cervigón_climate_2020;@giraldo_tree_2023]; drought timing can interact with this growth phenology especially.  -->
<!-- Sean::: don't think the above is true universally. Laura Boeschoten and Erick Esteban have separate papers well in prep showing some species demonstarating seasonal growth and others not -->
In tropical forests with dry seasons unfavorable to growth, tree drought responses may vary dramatically depending on the timing of drought relative to seasonal cycles of climate, photosynthesis, and woody growth [@garcía-cervigón_climate_2020]. <!---- agreed but see above ref I put in to [@albert_ecology_letters_cryptic_phenology]. It seems like these arguments are a bit redundant. Or meandering. ---->
<!--(Here we distinguish dry season from drought; dry season represents hot/dry conditions that are periodic, annual and predictable, droughts are unpredictable eith anomalous hot, dry conditions relative to seasonal expectations.) -->
Analyses to date have found that interannual variation in tropical tree growth <!--decreases with temperature[@zuidema_tropical_2022;@clark_annual_2010;@clark_annual_2021; @vlam_temperature_2014] and -->is particularly sensitive to dry season precipitation [@zuidema_tropical_2022;@clark_annual_2010;@clark_annual_2021]; however, these analyses represent a very limited set of tropical tree species and climates. 
It remains unknown how diverse assemblages of tropical tree species <!--in tropical dry forests -->respond to different types of droughts, including those that intensify or conflict with regular seasonal cycles<!-- of wet and dry seasons-->. 
<!--
Seasonally dry tropical forest climates are often close to thermal and water stress tolerance thresholds for many species therein, making these forests vulnerable to drought [@allen_will_2017] and leading to stand-level patterns of drought-associated water stress [@bauman_tropical_2022; @bennett_sensitivity_2023; @aguirre-gutiérrez_drier_2019]. <!--the last phrase of this sentence and how it relates to Bennett et al. is unclear to me.
Trees in seasonally dry tropical forests exhibit a mix of strategies - along a spectrum of conservative to acquisitive water-use - that support survival and growth despite strong seasonal hot/dry stress.
These strategies might confer differential advantages under drought conditions. -->

@@

<!-- Sean::: mention rooting depth? [@chitra-tarak: root of the drought and chitra-tarak: new_phyt 2022] -->
Tree species adaptations to drought vary widely in type and strength, often leading to classification of species as "drought tolerant" or "drought sensitive" [@guillemot_small_2022; @vico_how_2017a; @oliveira_linking_2021].
In tropical seasonal forests in particular, trees have evolved to maximise growth, reproduction, and survival under dry seasons, with strategies ranging along a spectrum from hydraulic safety to efficiency, leading to tolerance or avoidance of dry conditions [@gonzález-m._diverging_2021].
Dry-season deciduousness represents an extreme "avoidance" strategy along this continuum, where species lose leaves during the dry season to avoid foliar water loss [@desouza_drought_2020]. 
Other strategies include deep roots (hydraulic safety, to access deeper water in the dry season), more negative leaf water potential at turgor loss point ($\pi_{tlp}$, hydraulic safety, to maintain gas exchange and photosynthesis under hot and dry conditions), and large vessels (hydraulic efficiency)<!--but this may put them at risk of embolism--> [@gonzález-m._diverging_2021;@chitra-tarak_hydraulicallyvulnerable_2021]. 
Under drought conditions, some of these strategies can lend drought resistance while others can amplify drought sensitivity, especially under severe droughts [@kunert_leaf_2021; @chitra-tarak_hydraulicallyvulnerable_2021; @gonzález-m._diverging_2021]. 
<!-- This spectrum of species strategies in seasonally dry forests for demographic stability under hot dry seasons might have distinct consequences for sensitivity, based on drought characteristics.  -->
<!--Some theoretical evidence that the relative advantage of evergreen and deciduous strategies might differ with changing precipitation regimes [@vico_how_2017a]--->
Moreover, besides being adaptively conserved, these strategies can also have developmental or ecophysiological controls.
For example, some species regulate negative $\pi_{tlp}$'s, leaf deciduousness, and root depth based on environmental cues, leading to intraspecific or interannual variation and potentially resulting in spatiotemporal variation in drought responses [@williams_deciduousness_2008; @hulshof_variation_2010].

@@

Our study site was the Huai Kha Khaeng ForestGEO plot (15.6324 N, 99.217 E), hereafter "HKK", in the Huai Kha Khaeng Wildlife Sanctuary, Uthai Thani, Thailand.
This is a 50-hectare plot in dry seasonal evergreen forest, with mean annual precipitation of 1400 mm, mean annual temperature of 23.5 C and a strong dry season from November to April [@anderson-teixeira_ctfsforestgeo_2015].
The region is periodically affected by droughts associated with the *El Niño* Southern Oscillation [ENSO, @räsänen_spatial_2016].
The dominant canopy species are _Hopea odorata_, _Vatica harmandiana_, _Dipterocarpus alatus_ and _Tetrameles nudiflora_ and lower layers of the canopy are dominated by sub-canopy evergreen and deciduous species like _Miliusa horsfieldii_ and _Polyalthia viridis_. 
<!-- For analysis, we focused on 30 of the most abundant species in the plot.    -->
<!-- Analysis of long-term tree-ring data shows that interannual variation tree growth at HKK is influenced by temperature and rainfall [@vlam_temperature_2014].   -->

@@

Using the late wet/early dry dendrometer band censuses, we calculated annualised growth for each year and tree. 
<!-- We first conducted quality control on the dataset, removing measurements that appeared likely to be data entry errors (misplaced decimals), potential misidentified bands (old band numbers that restarted after stopping) or misidentified individuals (individuals with conflicting metadata across censuses). -->
We first conducted quality control on window size measurements (see supplementary methods) and combined them with the diameter measurement at installation $(d_{1})$ to calculate diameter at each dendroband census using standard equations that uses the chord across the measurement window to estimate the arc, assuming a circular cross section[@detto_correcting_2023].
We used a direct measurement of $d_{1}$ at the time of installation if it was available; if not, we excluded window size measurements until the first available diameter measurement. 
We calculated annualised diameter increments for each individual $i$ for each year $t$ by finding the difference in diameter from the previous year ($inc_{i,t} = d_{i,t} - d_{i,t-1}$) and adjusting for the number of days ($tdif$) between measurements ($ainc_{i, t} = inc_{i,t} * \frac{tdif}{365}$).

<!-- To ensure that the data did not smooth over multiple years, we excluded increment measures spanning missing values for $d_{t-1}$. -->
<!-- We excluded large measurement outliers (> 3 standard deviations from the mean increment across all observations).  -->
We conducted further quality control on increments (see supplementary methods for details) to assemble the final dataset. 
Diameter tape measurements themselves are not precise enough for calculating drought sensitivity; error for growth is twice measurement error = 2 mm. 
In comparison, median annual growth across trees in HKK is ~1.4 mm and calculating deviation from annual growth needs sub-mm precision. 
However, we used tape measurements to flag likely errors in the dendrometer dataset. 
<!-- We calculated increments from annual tape measurements made on each dendrobanded tree at each census and removed dendroband measurements that had low agreement.  -->
Finally, we excluded trees with negative increments over the whole timeseries, and included only species that had at least 10 individuals, resulting in a final dataset of 1820 individuals of 30 species.

@@

Across all individuals, we calculated Topographic Wetness Index (TWI) for an expectation of the water availability across space based on the total upslope area and the slope at the location. 
We used a Digital Elevation Model at 5 m spatial grain for the area of the plot for fine resolution and ability to locate the trees on the layer, but this omits upslope area that is off the plot. 
Using the *whitebox* package in R [@wu_whitebox_2022; @lindsay_whitebox_2016a] we calculated flow accumulation at each cell and then calculate TWI as $log_{e}\frac{As}{tan(slope)}$ where $As$ is the total upslope area within the plot and $slope$ is the absolute value of slope calculated at the 5m grain.
<!-- As an alternate measure of water availability, we also calculated Topographic Position Index (TPI) based on the concavity/convexity of a location.  -->
For an alternate estimate of wetness (Topographic Position Index) and associated results, see supplementary material. 
We defined tree size as the diameter at breast height (calculated from dendrometer band window measurement) at the previous census. 
We centred and scaled TWI and DBH across the range of observed values. 
We estimated crown exposure for each tree and year using ground measurements of crown illumination index (CII), an ordinal variable from 1 to 5 (least to most exposed).

We assigned deciduousness as a species-level trait using data from @williams_deciduousness_2008. 
We defined deciduousness as the mean over trees of the proportion of crown leaf loss at maximum, with a score ranging from 0 to 4 where 0 is most evergreen and 4 is most deciduous. 
For each species, these were calculated from observations of individual trees that were assigned categorical scores, and averaged, allowing species to take continuous, decimal values in the 0-4 range.
Data were unavailable for _Alphonsea ventricosa_, an evergreen species, and so we manually assigned it a deciduousness score of 1. 
We also considered an alternative measure of deciduousness - the duration of time spent at maximum crown loss.
This was strongly correlated with the proportion of crown loss; we chose to use the proportion of crown loss because it was available for more species.  
<!-- At the species level, we also calculated the maximum size across all individuals in the plot as measured in the 2008 whole-plot census.
We calculated niche breadth as the standard deviation of TWI across all observed individuals of each species in the whole plot census in 2008.  -->

@@

We then used the species effects (intercept + random effect) from this model and tested their correlations with deciduousness using simple linear models.

@@

<!-- We then calculated predictions across all combinations of values from observed maximum to minimum of TWI and deciduousness within this dataset. -->

To further disentange the microenvironmental drivers of drought sensitivity, we used a Bayesian causal model to simultaneously estimate fits for:

$$
sensitivity \sim 1 + DBH + CII + TWI + (1+DBH+CII+TWI | Species)
$$ 
and 
$$
CII \sim DBH
$$

### 2025-06-19 doc/supplementary_information.Rmd
some more comments from HML

- Data cleaning steps
- Calculating DBH equation
- Corroboration with tape measurements
- error distribution

>>>>

<!-- - Data cleaning steps -->
For raw measurements of dendrometer band window size, we conducted QAQC steps as follows:
1. Removed potentially misidentified individuals. We removed individuals with conflicting metadata on tag or location across censuses. We did not remove individuals with conflicting species identification because these are often updated during the ForestGEO censuses. Therefore, we used the latest version of species identification for each tag shared by the PIs of the HKK plot.
2. Removed potential misidentified bands. Each new dendroband installed on a tree is numbered sequentially starting from 1. We removed any measurements made on bands old bands after a new band series had begun. 
3. Removing measurements that appeared likely to be data entry errors. We identified misplaced decimals by checking if the ratio between adjacent window size measurements were closer to 1 or 10 or 100. We removed measurements that had ratios closer to 10 or 100 than 1, assuming that these were likely misplaced decimals.

After calculating DBH from window size measurements and calculating annualised growth increments, we conducted further QAQC on these increments to create the final dataset:
1. We excluded large measurement outliers, defined as > 3 standard deviations from the mean increment across all observations. 
2. We excluded trees with negative increments or increments close to zero over the whole timeseries of observations. These low growing trees may be indicative of stress or mortality, and likely to bias analysis of interannual growth variation. 
3. We used concurrent tape measurements on dendrobanded trees to flag likely errors. We calculated annualised increments from annual tape measurements made on each dendrobanded tree at each census, using similar methods as described for dendrometer bands. For each tree and year, we calculated the degree of deviation of these measurements from each other as the Euclidean distance from the 1:1 line. We found that 90% of the increments were within 5 mm of of deviation, which we retained as a high confidence dataset. We excluded the remaining 10% from the analysis, although we were unable to ascertain whether the discrepancy was because of errors in tape or dendroband measurements.
